Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-25 survey/survey.tex
lots of additions to sec 3

\documentclass{article}
\usepackage[utf8x]{inputenc}
\usepackage{agda}
\usepackage[nocenter]{qtree}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{listings}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usetikzlibrary{quotes,fit,positioning}
\tikzset{>=latex}

@@

@@

>>>>

\documentclass{article}
\usepackage[utf8x]{inputenc}
\usepackage{agda}
\usepackage[nocenter]{qtree}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{listings}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{proof}
\usetikzlibrary{quotes,fit,positioning}
\tikzset{>=latex}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\newcommand{\inkscape}[2][1.5]{
\begin{center}
\scalebox{#1}{
\includegraphics{inkscape/#2}
}
\end{center}
}

@@

\begin{definition}{Primitive operators of ~\ensuremath{\Pi } }
\label{def:primitives-langRev}

\[\begin{array}{rrcll}
 \mathit{zeroe} :& 0 + b &\leftrightarrow & b &: \mathit{zeroi} \\
 \mathit{swap}^{+} :& b_1 + b_2 &\leftrightarrow & b_2 + b_1 &: \mathit{swap}^{+} \\
 \mathit{assocl}^{+} :& b_1 + (b_2 + b_3) &\leftrightarrow & (b_1 + b_2) + b_3 &: \mathit{assocr}^{+} \\
 \mathit{unite} :& 1 \times b &\leftrightarrow & b &: \mathit{uniti} \\
 \mathit{swap}^{\times } :& b_1 \times b_2 &\leftrightarrow & b_2 \times b_1 &: \mathit{swap}^{\times } \\
 \mathit{assocl}^{\times } :& b_1 \times (b_2 \times b_3) &\leftrightarrow & (b_1 \times b_2) \times b_3 &: \mathit{assocr}^{\times } \\
 \mathit{distrib }_{0} :& 0 \times b &\leftrightarrow & 0 &: \mathit{factor }_{0} \\
 \mathit{distrib } :& (b_1 + b_2) \times b_3 &\leftrightarrow & (b_1 \times b_3) + (b_2 \times b_3) &: \mathit{factor } \\
 \end{array}\]
%subcode source isomorphisms.tex:427

\end{definition}

Each line of this table is to be read as the definition of one or two
operators. For example, corresponding to the \textit{identity of
  \ensuremath{\times}} isomorphism (\ensuremath{1\times b\leftrightarrow b}) we have the two operators
\ensuremath{\mathit{unite}:1\times b\leftrightarrow b} (reading the isomorphism from left to right) and
\ensuremath{\mathit{uniti}:b\leftrightarrow 1\times b} (reading the isomorphism from right to
left). These operators are inverses of each other.  Each of the two
cases of commutativity defines one operator that is its own inverse.

Having defined primitive operators, we need some means of composing
them. We construct the composition combinators out of the closure
conditions for isomorphisms.

\begin{definition}{Composition in \ensuremath{\Pi } }
$$
\infer{ \mathit{id} : b \leftrightarrow b}{
	 ~
}
\quad
\infer{ \mathit{sym} ~c : b_2 \leftrightarrow b_1}{
	 c : b_1 \leftrightarrow b_2
}
\quad
\infer{ c_1\fatsemi c_2 : b_1 \leftrightarrow b_3}{
	 c_1 : b_1 \leftrightarrow b_2
	&
	 c_2 : b_2 \leftrightarrow b_3
}
$$
$$
\infer{ c_1 + c_2 : b_1 + b_2 \leftrightarrow b_3 + b_4}{
	 c_1 : b_1 \leftrightarrow b_3
	&
	 c_2 : b_2 \leftrightarrow b_4
}
\quad
\infer{ c_1 \times c_2 : b_1 \times b_2 \leftrightarrow b_3 \times b_4}{
	 c_1 : b_1 \leftrightarrow b_3
	&
	 c_2 : b_2 \leftrightarrow b_4
}
$$
%subcode source isomorphisms.tex:462
\end{definition}

Thus we have program constructs that witness reflexivity \ensuremath{\mathit{id}},
symmetry \ensuremath{\mathit{sym}}, and transitivity~\ensuremath{\fatsemi} and two parallel
composition combinators, one for sums~\ensuremath{+} and one for
pairs~\ensuremath{\times}.

\begin{definition}{(Syntax of \ensuremath{\Pi })}
\label{def:langRev}
We collect our types, values, and combinators, to get the full language
definition.
\[\begin{array}{rclr}
 \mathit{value} ~\mathit{types}, b &::=& 0 ~|~ 1 ~|~ b+b ~|~ b\times b \\
 \mathit{values}, v &::=& () ~|~ \mathit{left} ~v ~|~ \mathit{right} ~v ~|~ (v,v) \\
 \\
 \mathit{combinator} ~\mathit{types}, t &::=& b \leftrightarrow b \\
 \mathit{isomorphisms}, \mathit{iso} &::=& \mathit{zeroe} ~|~ \mathit{zeroi} ~|~ \mathit{swap}^{+} ~|~ \mathit{assocl}^{+} ~|~ \mathit{assocr}^{+} \\
 & ~|~ & \mathit{unite} ~|~ \mathit{uniti} ~|~ \mathit{swap}^{\times } ~|~ \mathit{assocl}^{\times } ~|~ \mathit{assocr}^{\times } \\
 & ~|~ & \mathit{distrib }_{0} ~|~ \mathit{factor }_{0} ~|~ \mathit{distrib } ~|~ \mathit{factor } \\
 \mathit{combinators}, c &::=& \mathit{iso} ~|~ \mathit{id} ~|~ \mathit{sym} ~c ~|~ c \fatsemi c ~|~ c + c ~|~ c \times c \\
 \end{array}\]
%subcode source isomorphisms.tex:483
\end{definition}

\begin{definition}{Operational Semantics for \ensuremath{\Pi }. }
\label{def:operational-langRev}
Given a program \ensuremath{c : b_1 \leftrightarrow b_2} in \ensuremath{\Pi }, we can run it by supplying
it with a value \ensuremath{ v_1 : b_1}. The evaluation rules \ensuremath{c ~v_1 \mapsto v_2} are given below.

Additive fragment:

\[\begin{array}{rlcl}
 \mathit{zeroe} & (\mathit{right} ~v) &\mapsto & v \\
 \mathit{zeroi} & v &\mapsto & \mathit{right} ~v \\
 \mathit{swap}^{+} & (\mathit{left} ~v) &\mapsto & \mathit{right} ~v \\
 \mathit{swap}^{+} & (\mathit{right} ~v) &\mapsto & \mathit{left} ~v \\
 \mathit{assocl}^{+} & (\mathit{left} ~v_1) &\mapsto & \mathit{left} ~(\mathit{left} ~v_1) \\
 \mathit{assocl}^{+} & (\mathit{right} ~(\mathit{left} ~v_2)) &\mapsto & \mathit{left} ~(\mathit{right} ~v_2) \\
 \mathit{assocl}^{+} & (\mathit{right} ~(\mathit{right} ~v_3)) &\mapsto & \mathit{right} ~v_3 \\
 \mathit{assocr}^{+} & (\mathit{left} ~(\mathit{left} ~v_1)) &\mapsto & \mathit{left} ~v_1 \\
 \mathit{assocr}^{+} & (\mathit{left} ~(\mathit{right} ~v_2)) &\mapsto & \mathit{right} ~(\mathit{left} ~v_2) \\
 \mathit{assocr}^{+} & (\mathit{right} ~v_3) &\mapsto & \mathit{right} ~(\mathit{right} ~v_3) \\
 \end{array}\]
%subcode source isomorphisms.tex:504

Multiplicative fragment:

\[\begin{array}{rlcl}
 \mathit{unite} & ((), v) &\mapsto & v \\
 \mathit{uniti} & v &\mapsto & ((), v) \\
 \mathit{swap}^{\times } & (v_1, v_2) &\mapsto & (v_2, v_1) \\
 \mathit{assocl}^{\times } & (v_1, (v_2, v_3)) &\mapsto & ((v_1, v_2), v_3) \\
 \mathit{assocr}^{\times } & ((v_1, v_2), v_3) &\mapsto & (v_1, (v_2, v_3)) \\
 \end{array}\]
%subcode source isomorphisms.tex:514

Distributivity and factoring:

\[\begin{array}{rlcl}
 \mathit{distrib } & (\mathit{left} ~v_1, v_3) &\mapsto & \mathit{left} ~(v_1, v_3) \\
 \mathit{distrib } & (\mathit{right} ~v_2, v_3) &\mapsto & \mathit{right} ~(v_2, v_3) \\
 \mathit{factor } & (\mathit{left} ~(v_1, v_3)) &\mapsto & (\mathit{left} ~v_1, v_3) \\
 \mathit{factor } & (\mathit{right} ~(v_2, v_3)) &\mapsto & (\mathit{right} ~v_2, v_3) \\
 \end{array}\]
%subcode source isomorphisms.tex:523

The evaluation rules of the composition combinators are given below:

$$
\infer{ \mathit{id} ~v \mapsto v}{
	 ~
}
\quad
\infer{ (\mathit{sym} ~c) ~v_1 \mapsto v_2}{
	 c{\mathit{^{\dagger} }} v_1 \mapsto v_2
}
\quad
\infer{ (c_1\fatsemi c_2) ~v_1 \mapsto v_2}{
	 c_1 ~v_1 \mapsto v
	&
	 c_2 ~v \mapsto v_2
}
$$
$$
\infer{ (c_1 + c_2) ~(\mathit{left} ~v_1) \mapsto \mathit{left} ~v_2}{
	 c_1 ~v_1 \mapsto v_2
}
\quad
\infer{ (c_1 + c_2) ~(\mathit{right} ~v_1) \mapsto \mathit{right} ~v_2}{
	 c_2 ~v_1 \mapsto v_2
}
$$
$$
\infer{ (c_1 \times c_2) ~(v_1, v_2) \mapsto (v_3, v_4)}{
	 c_1 ~v_1 \mapsto v_3
	&
	 c_2 ~v_2 \mapsto v_4
}
$$
%subcode source isomorphisms.tex:546

\end{definition}

Since there are no values that have the type \ensuremath{0}, the reductions for
the combinators \ensuremath{\mathit{zeroe}}, \ensuremath{\mathit{zeroi}}, \ensuremath{\mathit{distrib }_{0}} and \ensuremath{\mathit{factor }_{0}}
omit the impossible cases. Also, the semantics of \ensuremath{\mathit{sym} ~c} are
defined in terms of the adjoint of \ensuremath{c} (denoted by \ensuremath{c^\mathit{^{\dagger} }}),
which we formalize below.

\begin{definition}[Adjoints, \ensuremath{c^\mathit{^{\dagger} }}]
 The adjoint of a combinator \ensuremath{c} is defined by \ensuremath{c^\mathit{^{\dagger} }} as follows:

  \begin{itemize}
  \item For primitive isomophisms \ensuremath{c}, \ensuremath{c^\mathit{^{\dagger} }} is given by its
    isomorphic inverse operator from the table in Definition
    \ref{def:primitives-langRev}.

  \item \ensuremath{(c_1 \times c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } \times c_2^\mathit{^{\dagger} }}

  \item \ensuremath{(c_1 + c_2)^\mathit{^{\dagger} } = c_1^\mathit{^{\dagger} } + c_2^\mathit{^{\dagger} }}

  \item \ensuremath{(c_1\fatsemi c_2)^\mathit{^{\dagger} } = c_2^\mathit{^{\dagger} } \fatsemi c_1^\mathit{^{\dagger} }}. (Note that the
    order of combinators has been reversed).

  \end{itemize}

\end{definition}

\begin{definition}[Observational Equivalence. \ensuremath{c_1=c_2}]
\label{def:pi-eqv}
  Combinators \ensuremath{c_1} and \ensuremath{c_2}, where \ensuremath{c_1 : b_1 \leftrightarrow b_2} and
  \ensuremath{c_2:b_1\leftrightarrow b_2}, are observationally equivalent,\ensuremath{c_1=c_2}, iff

  \ensuremath{\forall  ~v_1:b_1, v_2 : b_2. ~~ c_1 ~v_1 \mapsto v_2 \Longleftrightarrow c_2 ~v_1 \mapsto v_2}.
\end{definition}

\begin{proposition}{Canonical Isomorphism.}
  For any type \ensuremath{b} there exists the isomorphism \ensuremath{b \leftrightarrow b^{|b|}}.
  \begin{proof}
    The fact that such an isomorphism exists is evident from the
    definition of size and what it means for two types to be
    isomorphic. While many equivalent constructions are possible
    for any type \ensuremath{b}, one such construction is given by \ensuremath{\lbrack\!\lbrack b\rbrack\!\rbrack}:

\[\begin{array}{rclr}
 \lbrack\!\lbrack 0 \rbrack\!\rbrack & = & \mathit{id} \\
 \lbrack\!\lbrack 1 \rbrack\!\rbrack & = & \mathit{id} \\
 \lbrack\!\lbrack 1+b \rbrack\!\rbrack & = & \mathit{id} + \lbrack\!\lbrack b \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1+b_2)+b_3 \rbrack\!\rbrack & = & \mathit{assocr}^{+} \fatsemi \lbrack\!\lbrack b_1+(b_2+b_3) \rbrack\!\rbrack \\
 \lbrack\!\lbrack b_1+b_2 \rbrack\!\rbrack & = & ( \lbrack\!\lbrack b_1 \rbrack\!\rbrack + \mathit{id}) \fatsemi \lbrack\!\lbrack b_1^{|b_1|} + b_2 \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1\times b_2)\times b_3 \rbrack\!\rbrack & = & \mathit{assocr}^{\times } \fatsemi \lbrack\!\lbrack b_1\times (b_2\times b_3) \rbrack\!\rbrack \\
 \lbrack\!\lbrack (b_1+b_2)\times b_3 \rbrack\!\rbrack & = & \mathit{distrib } \fatsemi \lbrack\!\lbrack b_1\times b_3+b_2\times b_3 \rbrack\!\rbrack \\
 \end{array}\]
%subcode source isomorphisms.tex:598

  \end{proof}
\end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{A Graphical Language for \ensuremath{\Pi } }

Combinators of \ensuremath{\Pi } can be written in terms of the operators
described in definition \ref{def:langRev} or via a graphical language
similar in spirit to those developed for Geometry of Interaction
\cite{DBLP:conf/popl/Mackie95} and string diagrams for category theory
\cite{selinger-graphical}. Modulo some conventions and shorthand we
describe here, the wiring diagrams are equivalent to the operator
based (syntactic) description of programs.  \ensuremath{\Pi } combinators
expressed in this graphical language look like ``wiring
diagrams''. Values take the form of ``particles'' that flow along the
wires. Computation is expressed by the flow of particles.

\begin{itemize}
\item
The simplest sort of diagram is the \ensuremath{\mathit{id} : b \leftrightarrow b} combinator which
is simply represented as a wire labeled by its type \ensuremath{b}. In more
complex diagrams, if the type of a wire is obvious from the context,
it may be omitted.

\inkscape{b-wire.pdf}

\noindent
Values flow from left to right in the graphical language of
\ensuremath{\Pi }.  When tracing a computation, one might imagine a value
\ensuremath{v} of type \ensuremath{b} on the wire, as shown below.

\inkscape{b-wire-value.pdf}

\item
The product type \ensuremath{b_1\times b_2} may be represented both as one wire
labeled \ensuremath{b_1\times b_2} or by two parallel wires labeled \ensuremath{b_1} and
\ensuremath{b_2}. Both representations may be used interchangeably.

\begin{multicols}{2}
\inkscape{product-one-wire.pdf}
\inkscape{product-two-wires.pdf}
\end{multicols}

When tracing execution using particles, one should think of one
particle on each wire or alternatively as in folklore in the
literature on monoidal categories as a ``wave.''

\begin{multicols}{2}
\inkscape{product-one-wire-value.pdf}
\inkscape{product-two-wires-value.pdf}
\end{multicols}

\item
Sum types may similarly be represented using using parallel wires with
a \ensuremath{+} operator between them.

\begin{multicols}{2}
\inkscape{sum-one-wire.pdf}
\inkscape{sum-two-wires.pdf}
\end{multicols}

\noindent
When tracing the execution of \ensuremath{b_1+b_2} represented by one
wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  When tracing
the execution of two additive wires, a value can reside on only one of
the two wires.

\begin{multicols}{2}
\inkscape{sum-two-wires-left-value.pdf}
\inkscape{sum-two-wires-right-value.pdf}
\end{multicols}

\item
When representing complex types like \ensuremath{(b_1\times b_2)+b_3} some visual
grouping of the wires may be done to aid readability. The exact type
however will always be clarified by the context of the diagram.

\inkscape{complex-type-ambiguity.pdf}

\item
Associativity is entirely skipped in the graphical language. Hence
three parallel wires may be inferred as \ensuremath{b_1\times (b_2\times b_3)} or
\ensuremath{(b_1\times b_2)\times b_3}, based on the context. This is much like handling of
associativity in the graphical representations of categories.

\inkscape{assoc.pdf}

\item
Commutativity is represented by crisscrossing wires.

\begin{multicols}{2}
\inkscape{swap_times.pdf}
\inkscape{swap_plus.pdf}
\end{multicols}

\noindent
When tracing the execution of \ensuremath{b_1+b_2} represented by one
wire, one can think of a value of the form \ensuremath{\mathit{left} ~v_1} or \ensuremath{\mathit{right} ~v_2}
as flowing on the wire, where \ensuremath{v_1:b_1} and \ensuremath{v_2:b_2}.  By visually
tracking the flow of particles on the wires, one can verify that the
expected types for commutativity are satisfied.

\begin{multicols}{2}
\inkscape{swap_times_value.pdf}
\inkscape{swap_plus_value.pdf}
\end{multicols}

\item
The morphisms that witness that \ensuremath{0} and \ensuremath{1} are the additive and
  multiplicative units are represented as shown below. Note that since there
  is no value of type 0, there can be no particle on a wire of type \ensuremath{0}.
  Also since the monoidal units can be freely introduced and eliminated, in
  many diagrams they are omitted and dealt with explicitly only when they are
  of special interest.

\begin{multicols}{2}
\inkscape{uniti.pdf}
\inkscape{unite.pdf}
\end{multicols}

\begin{multicols}{2}
\inkscape{zeroi.pdf}
\inkscape{zeroe.pdf}
\end{multicols}

\item
Distributivity and factoring are represented using the dual
boxes shown below:

\begin{multicols}{2}
\inkscape{dist.pdf}
\inkscape{factor.pdf}
\end{multicols}

Distributivity and factoring are interesting because they represent
interactions between sum and pair types. Distributivity should
essentially be thought of as a multiplexer that redirects the flow of
\ensuremath{v:b} depending on what value inhabits the type \ensuremath{b_1+b_2}, as shown
below.

\begin{multicols}{2}
\inkscape{dist-wire-value1.pdf}
\inkscape{dist-wire-value2.pdf}
\end{multicols}

\noindent
Factoring is the corresponding adjoint operation.

\begin{multicols}{2}
\inkscape{factor-wire-value1.pdf}
\inkscape{factor-wire-value2.pdf}
\end{multicols}

\item Combinators can be composed in series (\ensuremath{c_1 \fatsemi c_2}) or
  parallel. Sequential (series) composition corresponds to connecting
  the output of one combinator to the input of the next.

\inkscape{c1c2_seq.pdf}

There are two forms of parallel composition -- combinators
  can be combined additively \ensuremath{c_1+ c_2} (shown on the left) or
  multiplicatively \ensuremath{c_1\times c_2} (shown on the right).

\begin{multicols}{2}
\inkscape{c1c2_par_sum.pdf}
\inkscape{c1c2_par_times.pdf}
\end{multicols}

\end{itemize}

\noindent
\textit{Example.} As an example consider the wiring diagram
of the combinator \ensuremath{c} below:

\ensuremath{c : b \times (1+1) \leftrightarrow b + b}

\ensuremath{c = \mathit{swap}^{\times } \fatsemi \mathit{distrib } \fatsemi (\mathit{unite} + \mathit{unite})}

\begin{center}
\scalebox{1.5}{
%%subcode-line{pdfimage}[diagrams/example1-crop.pdf]
\includegraphics{inkscape/example1.pdf}
}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Expressing Programs with \ensuremath{\Pi } }
\label{sec:langRev-examples}

In this section we introduce programming in \ensuremath{\Pi } through
several simple examples. We subsequently discuss the expressiveness of
the language and discuss various properties.

\label{examples}

\paragraph*{Booleans}
Let us start with encoding booleans. We use the type \ensuremath{1+1} to
represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
\ensuremath{\mathit{right}~()} representing \ensuremath{\mathit{false}}.
Boolean negation is straightforward to define:

\ensuremath{\mathit{not} : \mathit{bool} \leftrightarrow \mathit{bool}}

\ensuremath{\mathit{not} = \mathit{swap}^{+}}

\noindent
It is easy to verify that \ensuremath{\mathit{not}} changes \ensuremath{\mathit{true}} to \ensuremath{\mathit{false}} and
vice versa.

\paragraph*{Bit Vectors.}
We can represent $n$-bit words using an n-ary product of
\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
using the type \ensuremath{\mathit{bool}\times (\mathit{bool}\times \mathit{bool})}.  We can perform various
operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
three \ensuremath{\mathit{not}} operations:

\ensuremath{\mathit{not}_{\mathit{word}_3} :: \mathit{word}_3 \leftrightarrow \mathit{word}_3}

\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not} \times (\mathit{not} \times \mathit{not})}

\noindent We can express a 3-bit word reversal operation as follows:

\ensuremath{\mathit{reverse} : \mathit{word}_3 \leftrightarrow \mathit{word}_3}

\ensuremath{\mathit{reverse} = \mathit{swap}^{\times } \fatsemi (\mathit{swap}^{\times } \times \mathit{id})~ \fatsemi \mathit{assocr}^{\times }}

\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
many others we will see in this thesis, is formed by sequentially
composing several simpler combinators. Instead of presenting the
operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
for presentation reasons) to flatten the tree into a sequence of
reductions as caused by each component. Such a sequence of reductions
is given below:
\[\begin{array}{rlr}
 & (v_1, (v_2, v_3)) \\
 \mathit{swap}^{\times } & ((v_2, v_3), v_1) \\
 \mathit{swap}^{\times } \times \mathit{id} & ((v_3, v_2), v_1) \\
 \mathit{assocr}^{\times } & (v_3, (v_2, v_1)) \\
 \end{array}\]
%subcode source isomorphisms.tex:979

\noindent On the first line is the initial value. On each subsequent
line is a fragment of the \ensuremath{\mathit{reverse}} combinator and the value that
results from applying this combinator to the value on the previous
line. For example, \ensuremath{\mathit{swap}^{\times }} transforms \ensuremath{(v_1, (v_2, v_3))} to
\ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
the bits in reverse order.

We can also draw out the graphical representation of the 3-bit reverse
combinator. In the graphical representation, it is clear that the
combinator achieves the required shuffling.

\inkscape{reverse-3-bit.pdf}

\paragraph*{Conditionals.}
Even though \ensuremath{\Pi } lacks conditional expressions, they are
expressible using the distributivity and factoring laws. The
diagrammatic representation of \ensuremath{\mathit{distrib }} shows that it redirects the flow
of a value \ensuremath{v:b} based on the value of another one of type
\ensuremath{b_1+b_2}. If we choose \ensuremath{1+1} to be \ensuremath{\mathit{bool}} and apply either
\ensuremath{c_1:b\leftrightarrow b} or \ensuremath{c_2:b\leftrightarrow b} to the value \ensuremath{v}, then we essentially
have an `if' expression.

\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool} \times b \leftrightarrow \mathit{bool} \times b}

\ensuremath{\mathit{if}_{c_1,c_2} = \mathit{distrib } \fatsemi ((\mathit{id} \times c_1) + (\mathit{id} \times c_2)) \fatsemi \mathit{factor }}

\inkscape{if-c1-c2.pdf}

The diagram above shows the input value of type \ensuremath{(1+1)\times b} processed
by the distribute operator \ensuremath{\mathit{distrib }}, which converts it into a value of
type \ensuremath{(1\times b)+(1\times b)}. In the \ensuremath{\mathit{left}} branch, which corresponds to the
case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was \ensuremath{\mathit{left} ~()}),
the combinator~\ensuremath{c_1} is applied to the value of type~\ensuremath{b}. The right
branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
the value of type \ensuremath{b} through the combinator \ensuremath{c_2}.

\paragraph*{Logic Gates}
There are several universal primitives for conventional (irreversible)
hardware circuits, such as \ensuremath{\mathit{nand}} and \ensuremath{\mathit{fanout}}. In the case
of reversible hardware circuits, the canonical universal primitive is
the Toffoli gate~\cite{Toffoli:1980}. The Toffoli gate takes three
boolean inputs: if the first two inputs are \ensuremath{\mathit{true}} then the third
bit is negated. In a traditional language, the Toffoli gate would be
most conveniently expressed as a conditional expression like:

\noindent
\ensuremath{ \mathit{toffoli}(v_1,v_2,v_3) = \mathit{if} ~(v_1 ~\mathit{and} ~v_2) ~\mathit{then} ~(v_1, v_2, \mathit{not}(v_3)) ~\mathit{else} ~(v_1, v_2, v_3)}

We will derive Toffoli gate in \ensuremath{\Pi } by first deriving a simpler
logic gate called \ensuremath{\mathit{cnot}}.  Consider a one-armed version, \ensuremath{\mathit{if}_c},
of the conditional derived above. If the \ensuremath{\mathit{bool}} is
\ensuremath{\mathit{true}}, the value of type \ensuremath{b} is modified by the operator \ensuremath{c}.

\begin{center}
\scalebox{1.5}{
%%subcode-line{pdfimage}[diagrams/if_c.pdf]
\includegraphics{inkscape/cnot.pdf}
}
\end{center}

By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool}\times \mathit{bool}\leftrightarrow \mathit{bool}\times \mathit{bool}} which negates its
second argument if the first argument is \ensuremath{\mathit{true}}. This gate
\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.

If we iterate this construction once more, the resulting combinator
\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool}\times (\mathit{bool}\times \mathit{bool})\leftrightarrow \mathit{bool}\times (\mathit{bool}\times \mathit{bool})}. The
resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
Toffoli gate.

\begin{center}
\scalebox{1.6}{
\includegraphics{inkscape/toffoli.pdf}
}
\end{center}
